Staged Abstract Interpreter

### 2019-09-29 oopsla19-paper/50_stageabs.tex
reorganize

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change.
We manually fuse the three inner transformers (@IdM@ is omitted) into a single
monad @RepSetReaderStateM[R, S, A]@, where @R@ is the type parameter for the
reader effects, and @S@ is the type for the state.  In our monad stack scheme,
@R@ and @S@ are both instantiated with the type @Cache@. Similar to the unstaged
version, the grounded result type is a pair of two staged values:
@Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ... }
\end{lstlisting}

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that where the evaluation @ev(e)@ is invoked are shifted from @ap_clo@ to
@close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we 
focus on discussing the closure representation and function
application for the staged version.

@@

Our staging schema works by unfolding the interpreter over the abstract syntax
tree of the input program. In practice, however, the staging schema would
suffer from code explosion when analyzing (specializing) large programs, which
increases compile time. If we generate next-stage programs running on the JVM,
such large generated programs would also incur GC overhead at runtime. In this
section, we present optimizations that largely mitigate these issues.
Implementing all of those optimizations do not need to change the generic
interpreter.

@@

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is unnecessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in 
administrative normal form (ANF) $\lambda$-calculus
\cite{Flanagan:1993:ECC:155090.155113}, we can use a selective caching
algorithm that does not generate caching code for atomic expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

We have implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of @eval@ %our abstract
is applied to the proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize a small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} presented for
optimizing Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering effort.

>>>>

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as the unstaged abstract interpreter,
but replace \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change.
We manually fuse the three inner transformers (@IdM@ is omitted) into a single staged
monad @RepSetReaderStateM[R, S, A]@, where @R@ is the type parameter for the
reader effects, and @S@ is the type parameter for the state.  In our monad stack scheme,
@R@ and @S@ are both instantiated with the type @Cache@. Similar to the unstaged
abstract semantics, the grounded result type is a pair of two staged values:
@Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ... }
\end{lstlisting}

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, it
is observed that where we invoke the evaluator @ev(e)@ is shifted
from @ap_clo@ to @close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we 
focus on discussing the closure representation and function
application for the staged version.

@@

Our staging schema works by unfolding the interpreter over the abstract syntax
tree of the input program. In practice, however, the staging schema would
suffer from code explosion when analyzing (specializing) large programs, which
increases compile time. If we generate next-stage programs running on the JVM,
such large generated programs would also incur GC overhead at runtime. In this
section, we present optimizations that largely mitigate these issues.
Implementing all of those optimizations does not need to change the generic
interpreter.

@@

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is unnecessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in 
administrative normal form (ANF) $\lambda$-calculus
\cite{Flanagan:1993:ECC:155090.155113}, we can use a selective caching
algorithm that does not generate caching code for atomic expressions.
This optimization is implemented by using @fix_select@ to dispatch expressions.
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

We have implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of @eval@ %our abstract
is applied to the proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize a small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587}.
\citet{Johnson:2013:OAA:2500365.2500604} presented this idea for optimizing
Abstract Abstract Machines. However, the generated abstract bytecode still
requires another small-step abstract machine to execute, which is an additional
engineering effort.

### 2019-09-29 oopsla19-paper/60_cases.tex
reorganize

We first revisit a similar technique called \textit{abstract compilation} (AC).
AC was introduced by \citet{Boucher:1996:ACN:647473.727587} as an
implementation technique for abstract interpretation-based static analyses.
Similar to the present paper, the idea of AC was inspired by partial evaluation:
the program is known statically and the interpretive overhead can be
eliminated. In AC, the compiled analysis can be represented by either text or
closures (higher-order functions). The minute difference is that the closures
can be executed immediately, while the textual programs need to be compiled and
loaded.

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, every function in the analyzer is refactored to return a
closure only taking an environment argument.  Therefore the overhead of
traversing the AST of input program is eliminated.

@@

\begin{figure}[h!]
\vspace{-1em}
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // omit code for checking caches
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\vspace{-0.5em}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\vspace{-2em}
\end{figure}

@@

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$-term to a next-stage Scala function. 
The specialization of the staged abstract interpreter to a
$\lambda$-term is also independent of other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

@@

The specialization (0-CFA and without store-widening) of @close@ to
@map@ generates the following next-stage Scala code, that is, the
function @x39@:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // list of argument values
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // preparing a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, initially x41
        val x59 = x58.union(x55);  x52 + (x54 -> x59) }
    ... }
\end{lstlisting}

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such a generated function is a modularly
reusable artifact, i.e., independent of its context.
The generated functions already eliminate the interpretation overhead and
monadic overhead via staging, and can therefore be considered as partial
summaries. However, we still need to run the generated function at the next
stage to obtain the analysis result.  At the next stage, we can provide
different abstract arguments, stores, and caches, depending on the context
where it is called.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to a modular-staged analyzer is
mechanized and for-free via staging. We do not need to design any
intermediate representation or data structures of summaries.  In some
scenarios, traditional summary-based analyses are able to carry more
information within a module, e.g., the summary of their internal
submodules. The staged approach presented here may still need to run
through the fixed-point for submodules.

>>>>

We first revisit a similar technique called \textit{abstract compilation} (AC).
AC was introduced by \citet{Boucher:1996:ACN:647473.727587} as an
implementation technique for abstract interpretation-based static analyses.
Similar to the present paper, the idea of AC was inspired by partial evaluation:
the program is known statically and the interpretive overhead can be
eliminated. In AC, the compiled analysis can be represented by either text or
closures (higher-order functions). The minute difference is that the closures
can be executed immediately, while the textual programs need to be compiled and
loaded.
Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, every function in the analyzer is refactored to return a
closure only taking an environment argument.  Therefore the overhead of
traversing the AST of input program is eliminated.

@@

\begin{figure}[h!]
\vspace{-1em}
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out, and checking caches
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // the grounded result
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\vspace{-0.5em}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\vspace{-2em}
\end{figure}

@@

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$-term to a next-stage Scala function, which is the
specialized analyzer to that $\lambda$-term.  The specialization of the staged
abstract interpreter to a $\lambda$-term is independent of other parts of
the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

@@

The specialization (0-CFA without store-widening) of @close@ to
@map@ generates the following next-stage Scala code, that is, the
function @x39@:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // list of argument values
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // preparing a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, initially x41
        val x59 = x58.union(x55);  x52 + (x54 -> x59) }
    ... }
\end{lstlisting}

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such a generated function is a modularly
reusable artifact, i.e., independent of its context.
The generated functions already eliminate the interpretation overhead and
monadic overhead via staging, and can therefore be considered as partial
summaries. However, we still need to run the generated function at the next
stage to obtain the analysis result.  At the next stage, we can provide
different abstract arguments, stores, and caches, depending on the context
of where @map@ is called.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to a modular-staged analyzer is
mechanized via staging. We do not need to design any
intermediate representation or data structure of summaries.  In some
scenarios, traditional summary-based analyses are able to carry more
information within a module, e.g., the summary of their internal
submodules. The staged approach presented here may still need to run
through the fixed-point for submodules.

### 2019-09-29 oopsla19-paper/70_eval.tex
reorganize

We use Scala 2.12.8 and Oracle JVM 1.8.0-191 \footnote{The options for
running the JVM are set to the following: \texttt{-Xms2G -Xmx8G -Xss1024M
-XX:MaxMetaspaceSize=2G - XX:ReservedCodeCacheSize=2048M}},
running on an Ubuntu 16.04 LTS (kernel 4.4.0) machine. All our
evaluations were performed on a machine with 4 Intel Xeon Platinum
8168 CPUs at 2.7GHz and 3 TB of RAM. Although the machine has 96 cores
(192 threads) in total, the abstract interpreters are single-threaded.
To minimize the effect of warm-up from the HotSpot JIT compiler,
all the experiments are executed for 20 times and we report the
statistical median value of the running times. We set a 5 minute
timeout.

@@

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on the two monovariant CFAs. The
\textit{unstaged} and \textit{staged} columns show the median time (excluding
pre-compilation) to finish the analysis. The column
$\frac{\text{unstaged}}{\text{staged}}$ shows the improvement. The dash @-@ in
some cells indicates timeout.  As expected, the staged version significantly
outperforms the unstaged version on all benchmarks.
%We observe that simple value domains used in the benchmark usually leads to
%better speed-ups, for example, the @kcfa-worst@ series only uses lambdas and
%booleans.
For 0-CFA without store-widening, we observe an average speedup of ~17 times;
for 0-CFA with store-widening, the average speedup is ~7 times.
\texttt{kcfa-worst-64} has the highest speedup among all the tests. Although it
has a comparably large size of AST, it has a rather simple structure which only
involves booleans, $\lambda$-terms, and applications. We conjecture that the simple
domain and structure allow the generated code to be efficiently optimized by the
JVM. However, on the corresponding 0-CFA-SW, we do not observe such a significant
speedup. Due to the complexity of generated code and next-stage running platform
(JVM), we leave the detailed empirical analysis of which static analysis can
benefit more from the specialization as future work.

>>>>

We use Scala 2.12.8 and Oracle JVM 1.8.0-191 \footnote{The options for
running the JVM are set to the following: \texttt{-Xms2G -Xmx8G -Xss1024M
-XX:MaxMetaspaceSize=2G - XX:ReservedCodeCacheSize=2048M}},
running on an Ubuntu 16.04 LTS machine. All our
evaluations were performed on a machine with 4 Intel Xeon Platinum
8168 CPUs at 2.7GHz and 3 TB of RAM. Although the machine has 96 cores in
total, the abstract interpreters are single-threaded.  To minimize the effect
of warm-up from the HotSpot JIT compiler, all the experiments are executed 20
times and we report the statistical median value of the running times. We set a
5 minute timeout.

@@

Figure~\ref{evaluation_result} shows the evaluation result comparing the
performance improved by staging on the two monovariant CFAs. The
\textit{unstaged} and \textit{staged} columns show the median time (excluding
pre-compilation) to finish the analysis. The column
$\frac{\text{unstaged}}{\text{staged}}$ shows the improvement. The dash @-@ in
some cells indicates timeout.  As expected, the staged version significantly
outperforms the unstaged version on all benchmarks.
%We observe that simple value domains used in the benchmark usually leads to
%better speed-ups, for example, the @kcfa-worst@ series only uses lambdas and
%booleans.
For 0-CFA without store-widening, we observe an average speedup of ~17 times;
for 0-CFA with store-widening, the average speedup is ~7 times.
The highest speedup among all the tests is observed on benchmark \texttt{kcfa-worst-64}.
Although the program size of \texttt{kcfa-worst-64} is large, it has a rather
simple structure which only involves booleans, $\lambda$-terms, and
applications. We conjecture that the simple domain and structure allow the
generated code to be efficiently optimized by the JVM. However, on the
corresponding 0-CFA-SW, we do not observe a significant speedup. Due to
the complexity of generated code and next-stage running platform (JVM), we
leave the detailed empirical analysis of which static analysis can benefit more
from the specialization as a future work.

### 2019-09-29 oopsla19-paper/abstract.tex
reorganize

It is well known that a staged interpreter is a compiler: specializing an
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking an interpreter to run on abstract domains produces a sound static
analysis. What happens when we combine these two ideas, and apply specialization to
an \emph{abstract} interpreter?

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
generic interpreter with type-level binding-time abstractions and
monadic abstractions. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in one of four modes: as an unstaged
concrete interpreter, a staged concrete interpreter, an unstaged abstract
interpreter, or a staged abstract interpreter. As an example of
\emph{abstraction without regret}, the overhead of these abstraction layers is
eliminated in the generated code after staging. We show that staging abstract
interpreters is practical and useful to optimize static analysis, all while
requiring less engineering effort and without compromising soundness. We conduct
three case studies, including a comparison with Boucher and Feeley's abstract
compilation, applications to various control-flow analyses, and a demonstration
for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.

>>>>

It is well known that a staged interpreter is a compiler: specializing an
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking an interpreter to run on abstract domains produces a sound static
analysis. What happens when we combine these two ideas, and apply specialization to
an abstract interpreter?

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
generic interpreter with type-level binding-time abstractions and
monadic abstractions. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in one of four modes: as an unstaged
concrete interpreter, a staged concrete interpreter, an unstaged abstract
interpreter, or a staged abstract interpreter. As an example of
abstraction without regret, the overhead of these abstraction layers is
eliminated in the generated code after staging. We show that staging abstract
interpreters is practical and useful to optimize static analysis, all while
requiring less engineering effort and without compromising soundness. We conduct
three case studies, including a comparison with Boucher and Feeley's abstract
compilation, applications to various control-flow analyses, and a demonstration
for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.
